---
title: "giottoMulti: multi-sample container"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{giottoMulti: multi-sample container}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## What it is

`giottoMulti` is a container for **multiple `giotto` objects analyzed in a shared expression space**. Each child keeps its own spatial information (different physical / embedding spaces); shared analyses (joint normalization, dim reduction, NN graphs, clustering, integration) live on the parent.

This is distinct from `joinGiottoObjects()`, which merges multiple giottos into one — destructive, loses per-sample addressability. `giottoMulti` keeps the children intact and provides a non-destructive view layer on top.

## When to use it

Use `giottoMulti` when you have several spatial datasets (e.g. Visium slides, MERFISH samples) that share a feature panel and you want to:

* run joint expression-space analysis across them
* address each sample independently for per-sample operations (transforms, plotting, QC)
* iterate on subsets / filters without re-running integration each time
* keep the option of merging into a single `giotto` later via `joinGiottoObjects()`

## Construction

```{r}
library(GiottoClass)

# Two minimal samples, same feature panel
.mk <- function(ncell, nfeat) {
    m <- matrix(0, nrow = nfeat, ncol = ncell)
    rownames(m) <- paste0("f", seq_len(nfeat))
    colnames(m) <- paste0("c", seq_len(ncell))
    createGiottoObject(expression = m, verbose = FALSE)
}
g1 <- .mk(5, 4)
g2 <- .mk(3, 4)

mg <- createGiottoMulti(list(visium = g1, viz = g2))
mg
```

The `show` output surfaces:

* the children and their cell / feature counts
* the active scope (when narrower than all children)
* the current view: how many cells / features are visible vs total
* which joint shared slots are populated

## Global vs local IDs

Children's cell IDs are *local* (`c1`, `c2`, …). The multi namespaces them as **`sample::id`** globals via `@id_map`:

```{r}
spatIDs(mg)
spatIDs(mg, local = TRUE)
spatIDs(mg, object = "visium")
```

Feature IDs are passthrough by default — the feature panel is shared vocabulary across children.

## Joint expression: assembly then override

`getExpression(mg)` does the right thing depending on whether the joint slot is populated:

```{r}
# No joint slot populated yet — assembles naively from children
e <- getExpression(mg)
dim(e[])
colnames(e[])
```

This is the baseline view: per-child expression matrices concatenated with `sample::id` column names, features intersected. Useful for QC and inspection before any batch correction has been run.

Once you run integration (Harmony, scVI, Symphony, etc.) on the joint matrix and get a corrected one back, write it via `setExpression(mg, joint_matrix)` — that slot now becomes authoritative, and `getExpression(mg)` reads from it.

## Non-destructive subset

`subset()` narrows `@id_map` only — children are untouched, joint slots stay full. The view filter applies on every read.

```{r}
mg_qc <- subset(mg, cells = c("visium::c1", "visium::c2", "viz::c1"))
mg_qc

# Reads reflect the narrowed view
ncol(getExpression(mg_qc)[])

# Children intact — full data still there
length(spatIDs(mg_qc@objects$visium))
```

To widen back to the full view:

```{r}
mg_full <- rebuildMaps(mg_qc)
spatIDs(mg_full)
```

To commit to a subset and free the dropped joint data from memory:

```{r}
# populate something in a joint slot so compact has work to do
e2 <- getExpression(g1)
mat <- e2[]
colnames(mat) <- paste("visium", colnames(mat), sep = "::")
e2[] <- mat
mg_qc@expression <- list(cell = list(rna = list(raw = e2)))

mg_compact <- compact(mg_qc)
# joint @expression slot is now sized to the surviving view
ncol(mg_compact@expression$cell$rna$raw[])
```

`compact()` is one-way — once dropped, that data is gone. Use after committing to a cohort; keep things lazy via the view filter until then.

## Per-child operations

Spatial-domain accessors take an `object =` argument to route per child:

```{r}
# Spatial locations for one specific child
getSpatialLocations(mg, object = "visium")

# All active children — returns a named list
sl_all <- getSpatialLocations(mg)
names(sl_all)
```

These also apply the view filter by default. Pass `unfiltered = TRUE` to bypass:

```{r}
mg_sub <- subset(mg, cells = "visium::c1")

# default: narrowed by view
nrow(getSpatialLocations(mg_sub, object = "visium")$visium[])

# escape hatch: full child content
nrow(getSpatialLocations(mg_sub, object = "visium", unfiltered = TRUE)$visium[])
```

Per-child setters require an explicit `object =` target:

```{r, eval=FALSE}
setSpatialLocations(mg, x = my_locs, object = "visium")
setPolygonInfo(mg, x = my_polys, object = "viz")
```

## Active scope

`@active` controls which children per-child operations route to by default. Set it via `activeObjects<-` to scope subsequent calls:

```{r}
activeObjects(mg) <- "visium"
# now getX operations on the multi only touch "visium"
spatIDs(mg)

activeObjects(mg) <- NULL  # reset to all
spatIDs(mg)
```

## Single-object wrap

You can wrap a single `giotto` as a one-child `giottoMulti` to use the view layer (subset, compact) without committing to a destructive subset:

```{r}
mg_single <- as(g1, "giottoMulti")
mg_single

# Same lazy view layer applies
mg_narrowed <- subset(mg_single, cells = c("sample1::c1", "sample1::c3"))
spatIDs(mg_narrowed)
rebuildMaps(mg_narrowed)
```

## Realigning after manual edits

If you edit a child directly (e.g. by replacing it with `mg[["a"]] <- new_g` or by mutating a slot outside the approved op surface), the cached `@id_map` may be out of sync. `initialize(mg)` re-detects via the cached length signature and rebuilds when needed — cheap on every call, no-op when nothing changed.

`rebuildMaps(mg)` forces a rebuild even when signatures match — use this if you've made a same-length-different-content change to a child.

## Relationship to other containers

| | giottoMulti | `joinGiottoObjects` | Seurat v5 layers | SCE / SpatialExperiment |
|---|---|---|---|---|
| Per-sample addressability | ✅ | ❌ (merged) | partial | by `sample_id` filter |
| Per-sample spatial state | ✅ rich | merged with offsets | thin | thin |
| Non-destructive subset | ✅ via view filter | ❌ | ❌ | ❌ |
| Joint analysis layer | ✅ | inherent | via JoinLayers | inherent |

`giottoMulti` and `joinGiottoObjects()` are siblings, not nested: one preserves per-sample spaces, the other merges. Convert between them when needed.
