---
title: "giottoMulti: working with several samples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{giottoMulti: working with several samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    fig.width = 7,
    fig.height = 3.5
)
options("giotto.use_conda" = FALSE, "giotto.no_python_warn" = TRUE)
```

A `giottoMulti` holds several `giotto` objects — one per sample — and lets you
analyze them together without merging them. Each sample keeps its own spatial
data and coordinate frame. What the samples share, such as a joint expression
matrix, joint metadata or an integrated embedding, lives on the multi.

Use one when the samples share a feature panel and you want joint analysis
across them while still being able to address, plot and edit each sample on its
own. If the samples are really one specimen and you want a single coordinate
system, `joinGiottoObjects()` is the better fit; see the comparison at the end.

## Building one

`createGiottoMulti()` takes a named list of `giotto` objects. The names become
the sample names.

```{r, message = FALSE}
library(GiottoClass)

mg <- createGiottoMulti(list(
    tissue_a = GiottoData::loadGiottoMini("visium", verbose = FALSE),
    tissue_b = GiottoData::loadGiottoMini("visium", verbose = FALSE)
))
mg
```

The two samples here are copies of the same mini dataset, which keeps the
example small; everything below works the same with genuinely different
samples.

A multi behaves like a named list of its samples:

```{r}
names(mg)
length(mg)

g_a <- mg[["tissue_a"]]   # one sample, as a plain giotto
class(g_a)
```

## Cell IDs are qualified by sample

Two samples can contain the same cell ID, so the multi refers to cells as
`sample::cell_ID`:

```{r}
head(spatIDs(mg), 3)
head(spatIDs(mg, local = TRUE), 3)          # the IDs inside each sample
length(spatIDs(mg, object = "tissue_b"))    # one sample's cells
```

Feature IDs are shared across samples, so they are not qualified.

## Reading data

The getters work on a multi as they do on a `giotto`, with a `samples =`
argument to pick which samples to read.

**Tabular data** — cell metadata, expression, feature metadata — comes back as
one joint table across the samples. Cell metadata carries a `list_ID` column
naming each cell's sample:

```{r}
cm <- getCellMetadata(mg, output = "data.table")
dim(cm)
table(cm$list_ID)

expr <- getExpression(mg, output = "matrix")
dim(expr)

dim(getExpression(mg, output = "matrix", samples = "tissue_a"))
```

**Spatial data** — locations, polygons, points, images, spatial networks —
belongs to one sample each, so it comes back as a named list with one entry per
sample:

```{r}
locs <- getSpatialLocations(mg)
names(locs)

locs_a <- getSpatialLocations(mg, samples = "tissue_a")[["tissue_a"]]
locs_a
```

A small helper draws the spatial locations it is given, one panel per sample.
It is reused below:

```{r}
plot_samples <- function(locs, highlight = NULL, ...) {
    op <- par(mfrow = c(1, length(locs)), mar = c(2, 2, 2, 1))
    on.exit(par(op))
    for (nm in names(locs)) {
        xy <- locs[[nm]][]
        plot(xy$sdimx, xy$sdimy, pch = 16, cex = 0.6, asp = 1,
            col = if (is.null(highlight)) "grey40" else
                ifelse(xy$cell_ID %in% highlight[[nm]], "firebrick", "grey85"),
            main = nm, xlab = "", ylab = "", ...)
    }
}

plot_samples(getSpatialLocations(mg))
```

## Naming a set of samples: groups

A group is a name for several samples. Once registered, it works anywhere a
sample name does:

```{r}
gmultiGroup(mg, "pair") <- c("tissue_a", "tissue_b")
gmultiGroups(mg)

nrow(getCellMetadata(mg, output = "data.table", samples = "pair"))
```

Groups resolve when they are used, so a group keeps up as samples are added,
renamed or removed. A group and a sample cannot share a name.

## How samples line up: the mapping

A multi needs to know which content in each sample belongs together — which
spatial unit, which feature type, which expression matrix. That is the mapping,
and it is filled in automatically when the multi is built:

```{r}
gmultiMapping(mg, "values")
```

Each entry maps a name at the multi level to the name each sample uses. Here
both samples call their raw counts `"raw"`. If one sample used a different name,
say `"counts"`, you would say so once, and every joint read would use it:

```{r, eval = FALSE}
gmultiMapping(mg, "values", "raw") <- c(tissue_a = "raw", tissue_b = "counts")
```

An `NA` entry means a sample deliberately does not contribute to that handle.
A sample that is supposed to contribute but cannot is an error when the data is
read, naming the sample.

## Joint data and per-sample data

Joint tables are assembled from the samples the first time they are read.
Anything written to the multi goes to the joint level and leaves the samples
alone:

```{r}
ids <- spatIDs(mg)
mg <- addCellMetadata(mg,
    new_metadata = data.table::data.table(
        cell_ID = ids,
        batch = ifelse(startsWith(ids, "tissue_a"), "batch1", "batch2")),
    by_column = TRUE, column_cell_ID = "cell_ID")

table(getCellMetadata(mg, output = "data.table")$batch)
"batch" %in% names(getCellMetadata(mg[["tissue_a"]], output = "data.table"))
```

Spatial data has no joint level, so it cannot be written through the multi:

```{r, error = TRUE}
setSpatialLocations(mg, locs_a)
```

To change one sample, take it out, edit it, and put it back:

```{r, warning = FALSE}
g <- mg[["tissue_b"]]
g <- spatShift(g, dx = 500)
mg[["tissue_b"]] <- g

range(getSpatialLocations(mg, samples = "tissue_b")[[1]][]$sdimx)
```

## Narrowing

`subset()` narrows the multi. With `cells =` it keeps those cells in the joint
analysis; with `samples =` it keeps those samples, the same as `mg[...]`:

```{r}
mg_small <- subset(mg, cells = head(spatIDs(mg), 100))
length(spatIDs(mg_small))

names(subset(mg, samples = "tissue_a"))
names(mg["tissue_b"])
```

Narrowing by cell leaves each sample's own spatial data untouched, so the full
tissue is still there if you need it:

```{r}
length(spatIDs(mg_small[["tissue_a"]]))
```

`subset()` returns a new object; the original `mg` is unchanged.

## Views: narrowing without changing the object

A view records a narrowing under a name instead of applying it. Build one with
the usual verbs and a `view =` argument, then read through it with any getter:

```{r}
mg <- subset(mg, leiden_clus == 1, view = "cluster1")

nrow(getCellMetadata(mg, output = "data.table"))
nrow(getCellMetadata(mg, output = "data.table", view = "cluster1"))
```

The same view narrows spatial reads, per sample:

```{r}
c1 <- getSpatialLocations(mg, view = "cluster1")
plot_samples(getSpatialLocations(mg),
    highlight = lapply(c1, function(x) x[]$cell_ID))
```

A view can also select samples. With `samples =` and `view =`, `subset()`
records a sample step, and reads through the view skip the other samples
entirely. One call can record both a sample step and a filter:

```{r}
mg <- subset(mg, leiden_clus == 1, samples = "tissue_b", view = "b_cluster1")

names(getSpatialLocations(mg, view = "b_cluster1"))
table(getCellMetadata(mg, output = "data.table", view = "b_cluster1")$list_ID)
```

Views travel with the object and can be copied between objects. See
`vignette("view_and_space", package = "GiottoClass")` for crops, the other step
kinds, and how views are stored.

## Spaces: laying samples out

A space records coordinate transforms under a name. On a multi, `samples =`
says which samples a transform moves, which is how you build a layout. Here
`tissue_b` is moved beside `tissue_a` without changing either sample's data:

```{r, fig.height = 2.6}
mg <- spatShift(mg, dx = 4000, space = "side_by_side", samples = "tissue_b")

laid_out <- getSpatialLocations(mg, space = "side_by_side")
xy <- rbind(
    cbind(laid_out$tissue_a[], sample = "tissue_a"),
    cbind(laid_out$tissue_b[], sample = "tissue_b"))
plot(xy$sdimx, xy$sdimy, pch = 16, cex = 0.5, asp = 1,
    col = ifelse(xy$sample == "tissue_a", "steelblue", "darkorange"),
    xlab = "", ylab = "")
legend("topright", legend = c("tissue_a", "tissue_b"), pch = 16,
    col = c("steelblue", "darkorange"), bty = "n", cex = 0.8)

range(getSpatialLocations(mg, samples = "tissue_b")[[1]][]$sdimx)   # native
```

A space built this way gives each sample its own copy of the frame, so the
samples stay independent: a job over it is one job per sample. When the samples
should share one coordinate system — so that distances between them mean
something — declare a combined space first, naming its members:

```{r}
giottoSpace(mg, "atlas") <- combinedSpace(c("tissue_a", "tissue_b"))
mg <- spatShift(mg, dx = 4000, space = "atlas", samples = "tissue_b")
giottoSpace(mg, "atlas")
```

Views and spaces combine: the view picks the cells, the space places them.

```{r}
placed <- getSpatialLocations(mg, view = "cluster1", space = "side_by_side")
sapply(placed, function(x) c(cells = nrow(x[]), range(x[]$sdimx)))
```

## Analysis and plotting

The analysis functions in the Giotto package — normalization, dimension
reduction, clustering, integration — work on a multi's joint expression and
write their results to its joint slots, alongside the per-sample data.

Plotting lives in GiottoVisuals. Spatial plots draw one panel per sample and
take the same `samples =`, `view =` and `space =` arguments as the getters.
Non-spatial plots (UMAP, violin, dot plots) draw one plot across all samples;
color by `list_ID` to tell them apart:

```{r, eval = FALSE}
GiottoVisuals::spatPlot2D(mg, samples = "pair", view = "cluster1")
GiottoVisuals::spatPlot2D(mg, space = "atlas")
GiottoVisuals::plotUMAP(mg, cell_color = "list_ID")
```

## `giottoMulti` or `joinGiottoObjects()`?

Both put several samples into one workflow. They differ in what they keep.

| | `giottoMulti` | `joinGiottoObjects()` |
|---|---|---|
| Each sample addressable by name | yes: `samples =`, groups | no: merged into one object |
| Spatial data | stays with each sample, in its own frame | merged, with coordinate offsets |
| A sample on its own | `mg[["name"]]` is a full `giotto` | not recoverable |
| Laying samples out | recorded spaces, changeable later | fixed at join time |
| Joint analysis | joint slots on the multi | the one object |

Pick `giottoMulti` when sample identity matters through the analysis — QC,
normalization, integration, per-sample plots, cohort comparisons. Pick
`joinGiottoObjects()` when the samples really are one specimen and you want a
single coordinate system.

## See also

* `vignette("view_and_space", package = "GiottoClass")` — views and spaces in
  depth.
* `vignettes/articles/design_gmulti.Rmd` — why `giottoMulti` is built the way it
  is.
* `?createGiottoMulti`, `?gmultiGroup`, `?gmultiMapping`, `?"subset-giottoMulti"`.
